---
title: "Probability and Dispersion"
subtitle: "Exercise 3"
thanks: "This exercise is written by Justin Steil, with substantial revisions by Eric Robsky Huntley."
output: 
  pdf_document: 
  template: '../templates/syllabus.latex'
---

# Part 1: Probability  

A recent survey of the Muswell Hill neighborhood north of London revealed the following:
  
+ Total Units: the neighborhood has 1,200 apartments;
+ Parking: 25% of the apartments have no dedicated parking spaces;
+ Size: one-sixth of the apartments are one-bedroom (1BR), one-half are two-bedroom (2BR), and one-third are three-bedroom (3BR);
+ Access: 10% of the apartments are wheelchair-accessible;
+ Pets: 240 of the apartments allow pets.

Please answer the following questions based on your knowledge of probability! **Assume that all of these events are independent.**

a. How many 1BR apartments with dedicated parking would you expect in this neighborhood?

b. What percentage of the wheelchair-accessible apartments with parking that allow pets are 3BR?

c. What percentage of the 3BR apartments with parking that allow pets are wheelchair-accessible?

d. A “limited-mobility-friendly" (LMF) apartment is defined as an apartment that has either dedicated parking or is wheelchair-accessible, or both. How many LMF apartments are there in the neighborhood that are 2BR or larger?

## Part 2: Standard Deviation

```{r echo=FALSE, include=FALSE}
df <- readr::read_csv("boston_mhi.csv")
rows <- 10
```

Examine this table that shows the Median Household Income in the `r rows` most populated neighborhoods in Boston. Manually calculate the mean, median, and standard deviation of this sample. Please show the steps.

```{r, echo=FALSE, results="asis"}
 df |>
  dplyr::slice_max(Households, n = rows) |>
  knitr::kable("latex")
```

Source: Boston Planning & Development Authority. 2024. _[Boston in Context: Neighborhoods](https://www.bostonplans.org/getattachment/528be767-2cb3-4685-9a66-206cede8773b)_. Based on 2018-2022 American Community Survey.

# Part 3: Getting to Know Your Data

In this assignment, we will continue to Jeff Kaufman's Boston rents database, which we've modified to include the Boston neighborhood in which the listed units fall.

1. Produce five-number summaries of asking rents for 2-bedroom units in Dorchester, Roxbury, and the South End.
2. What percentage of units in Chinatown have rents that fall below the U.S. Department of Housing and Urban Development's 2024 fair market rent for 1-bedrooms in Boston? ($2,377)
3. What are the variance and standard deviation of rents for 1-bedroom units in Allston?
4. How many 1-bedroom units in Jamaica Plain are more than two standard deviations above the mean asking rent for 1-bedrooms in that neighborhood? (Hint: calculate the value of two standard deviations above the mean first, then use it to `filter()`!)
5. Produce a boxplot that depicts summaries of rent by bedroom count in Jamaica Plain.

