---
title: "Introduction to R"
author: "Eric Robsky Huntley"
date: "Spring 2024"
output:
  html_document:
    df_print: paged
  pdf_document:
    template: ../../templates/syllabus.latex
subtitle: Recitation 2
thanks: This is a substantial extension and rewrite of material originally developed
  by Sebastian Sandoval Olascoaga and subsequently revised by Ben Preis.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  hold=TRUE,
  warning = FALSE,
  message = FALSE,
  size = 'huge'
  )
```

# Where is R? (The Working Directory)

When you open `R`, it has a directory that it is working out of. We can confirm this by running the `getwd()` command (for 'get working directory').

```{r,echo=TRUE,eval=FALSE}
getwd()
```

This is the place it's going to look for any files that you may want to read, and where it's going to put any files that you write from your `R` process. Oftentimes, this will not be ideal for your purposes. If you create a new script in a folder, you might want that folder to serve as the place for `R` to start looking. 

To illustrate: say my working path is `/Users/ehuntley/Desktop/`---this is where your Desktop files are stored on your disk on a Mac. (The Windows equivalent is `C:\Users\username\Desktop`.) This means that I can reference a file on the desktop in my `R` environment by simply calling its name with no prefix. In other words, looking for a file using `data.csv` will be the same as looking for `/Users/ehuntley/Desktop/data.csv`. But then suppose that I have a folder on my desktop for all my QR assignments (let's call it `QR`). Within that directory, I have a `recitations` folder, an `exercises` folder, etc. Then, let's say within my `recitations` folder, I create a folder called `recitation_02`, which has a `data.csv` file in it. To locate this file when `R`'s current working directory is the desktop, I'd have to specify a location like this: `qr/recitations/recitation_02/data.csv`. It works, but it's a lot of text and a lot of typing. So we might want to __set the working directory__ to be `recitation_02` (meaning that `data.csv` can be located using only its name).

We can set the working directory to `recitation_02` like this:

```{r, eval=FALSE}
setwd("/Users/ehuntley/Desktop/qr/recitations/recitation_02")
```

This is a good thing to do at the very top of your file!

# Reading Data

For this recitation we are going to use American Community Survey (ACS) 5-year estimates of occupant tenure by housing unit. (In other words, renter- and owner-occupancy.) I've provided this data as a CSV. We can read the CSV into an `R` dataframe using _either_ the `read.csv()` function from base `R` or the `read_csv()` function from the tidyverse's `readr` library (which gets installed as a component of the `tidyverse`). For our purposes, they do the same thing, though the `read_csv()` function will read the CSV in a way that leads processing on it to be slightly faster.

```{r}
# These do the same thing!
df <- read.csv('data/lecture_02_demo.csv')
# Except this one _can_ make your code run a little faster...
library(readr)
df <- read_csv('data/lecture_02_demo.csv')
```

First, examine your dataframe. (I.e., click on it in the `RStudio` 'environment' pane, or run `View(df)`). You'll see that it contains percentage values for both renter- and owner-occupied housing units. These are both continuous variables.

Let's say, though that we were interested in identifying which tracts were majority renter. We can create a new column (`maj_renter`) that indicates this using a boolean (i.e., `TRUE`/`FALSE` value). We'll use the `mutate()` function from `dplyr` (which, again, is installed with the `tidyverse`). This function allows us to calculate a new column (or re-calculate an existing column) using the values of other columns. Let's do all of this using a single pipe---read through the annotations to see how this is working!

```{r}
library(dplyr)
# Read data
df <- read_csv('data/lecture_02_demo.csv') |>
  # `|>` passes results of read_csv to mutate
  dplyr::mutate(
    # Calculate a new column called `maj_renter`
    # which is the result of testing whether each
    # value of `pct_rent_occ` is greater than
    # 50.
    maj_renter = pct_rent_occ > 50
  )
```

Examine the results---note the new column you have to work with, and that it takes one of two values, indicating whether the truth test `pct_rent_occ > 50` was `TRUE` or `FALSE` for each row.

# Frequency Table

Now, let's determine how many tracts are majority renter and how many are not. This is known as a __frequency table__. To calculate this quickly, we'll use the `xtabs()` function, which comes with the base `R` `stats` package.

We pass in the dataframe to the `data` parameter of `xtabs()` and pass it a function (indicated by the `~` syntax) that indicates which variables we're summarizing by, separated by a `+` sign. To calculate a frequency table (which is only a single variable), we just pass the `maj_renter` variable.

```{r}
xtabs(~maj_renter, data=df)
```

# Cross Tabulation

But frequently, we want to calculate frequencies for multiple combinations of variables. For example, the number of tracts that are majority- or minority- renter for each county in Massachusetts (given in the `county` column). All we need to do is add the `county` column to our formula in the `xtabs` function, like so...

```{r}
xtabs(~county + maj_renter, data=df)
```

We can see that the relative number of tracts that are majority-renter varies widely across the state, though we see predictably higher rates in Suffolk County (Boston), and, to a lesser extent, Middlesex County (e.g., Cambridge, Somerville, Medford)  Hampden County (e.g., Springfield), and Worcester County (e.g., Worcester).

# Histograms

Cross-tabulations and frequency tables like those above have gotten us part of the way to investigating patterns in our data. But to really get a sense of the shape of our data's distribution, we can use histograms and kernel density graphs. Recall that a histogram is a visual representation of frequency distribution, constructed by counting observations falling into (usually) regularly-spaced intervals.

For example, we can create a histogram of the percentage of housing units that are renter occupied in MA census tracts using the `hist()` function.

```{r}
hist(df$pct_rent_occ, 
     # Add a title
     main="Histogram of % Renter Occupancy in Massachusetts",
     # Label the x-axis
     xlab="% Renter Occupancy",
     # Label the y-axis
     ylab = "Frequency",
     #Tell R how man breaks to have
     breaks=20) 
```
Note that we're dealing with what we'd refer to as a long right tail---there is a large range of values lying above the most frequently observed values. Try playing with the `breaks` parameter: a larger number will yield smaller breaks and therefore a less 'smoothed' representation of the data's distribution.

This is an __absolute frequency__, which is to say a measure of counts. If we wanted to generate a histogram depicting _density_ so that the area under the histogram equalled `1`, we would simply need to set the `freq` parameter equal to `FALSE`.

```{r}
hist(df$pct_rent_occ, 
     main="Histogram of % Renter Occupancy in Massachusetts",
     xlab="% Renter Occupancy",
     ylab = "Frequency",
     breaks=20,
     freq = FALSE) 
```

+ You try it! Swap out owner-occupancy. What do you notice about how the two shaped are related (and why might this be)?

# Kernel density

A different method of visualizing a variable distribution is a __kernel density__ plot. This is much like a histogram, except it visualizes the data continuously by averaging the densities around each potential value of the variable of interest. In other words: it's a smoothing function.

To generate a kernel density, we simply `plot()` the results of applying the `density()` function to a given variable.

```{r}
plot(density(df$pct_rent_occ), 
  xlab = "% Renter-Occupancy",
  ylab = "Density",
  main = "Kernel Density of % Renter-Occupancy in Massachusetts"
  )
```

Or, to do the same thing using pipes...

```{r,eval=FALSE}
df$pct_rent_occ |> 
  density() |>
  plot(
    xlab = "% Renter-Occupancy",
    ylab = "Density",
    main = "Kernel Density of % Renter-Occupancy in Massachusetts"
  )
```

That's all for today!