---
title: "Cross Tabulations & Summary Statistics"
author: "Chenab Navalkha and Eric Robsky Huntley"
date: "Spring 2024"
output: pdf_document
thanks: This is a substantial extension and rewrite of material originally developed
  by Sebastian Sandoval Olascoaga and subsequently revised by Ben Preis and Gokul
  Sampath.
subtitle: Recitation 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  hold=TRUE,
  warning = FALSE,
  message = FALSE,
  size = 'huge'
  )
```

# Set Working Directory


To start, we'll set the Working Directory for this week's recitation along the same lines as last week. You'll want to start by downloading the recitation folder (with its data) to the appropriate place on your computer. 

Once you have downloaded the recitation materials, use the `setwd()` function to specify the location of your working directory.


```{r, eval=FALSE}
setwd("/Users/chenabnavalkha/Desktop/QR/recitations/recitation_03/")
```


# Reading Data

For this recitation we are going to use data on Boston rents. The dataset we'll use is scraped from Padmapper.com, an online marketplace for rental units, and published by Jeff Kaufman. (You can see more of Jeff's analyses and find more information on the dataset here: https://www.jefftk.com/p/boston-rents-over-time-ii). 

The data saved in this week's recitation materials combines the scraped data on rental prices and number of bedrooms with a City of Boston data file from which we pulled in neighborhood information for each apartment listing. 

We've provided this data as a CSV. Read this CSV into an `R` dataframe using _either_ the `read.csv()` function from base `R` or the `read_csv()` function from the tidyverse's `readr` library. If you use `read_csv()`, remember to load the `tidyverse` package. 


```{r, eval=TRUE, include=TRUE}
library(tidyverse)
bos_rents <- read_csv("data/boston_rents.csv")
```

As a first step, examine your dataframe. To do this, you can either click on the dataframe name in the `RStudio` 'environment' pane, or run `View(df)`). 

```{r, eval=FALSE, include=TRUE}
View(bos_rents)
```


## Examining Variables of Interest

Before we begin making tables, it's good practice to look at just the variables of interest to make sure nothing is awry. 

Use the `unique` function from base R to examine the unique neighborhoods and number of bedrooms present in our rent data:

```{r, include=TRUE}
unique(bos_rents$neighborhood)
unique(bos_rents$bedrooms)
```

This shows us a list of the neighborhoods in our rent price data, as well as the range in number of bedrooms for the homes and apartment units included in the dataset. 


# Making tables and checking frequencies

How do these rental listings fall out by neighborhood? We can use the `cross tabulation` function `xtabs`to look at the number of observations (i.e. data points) by neighborhood. We will be using the xtable package again.

```{r, include=TRUE}
library(xtable)
xtabs(~neighborhood,data=bos_rents)
```


We can also compare two variables against each other. Let's create a table that shows us how many observations we have by number of bedrooms and neighborhood. 

```{r,echo=TRUE}
xtabs(~bedrooms+neighborhood, data=bos_rents)
```

# Frequencies and Distributions

Next, let's examine the frequency of rent prices by displaying a histogram that shows the absolute frequency (number of observations) on the y-axis and rent on the x-axis. 

Here, we're using the `pull()` function (from the `tidyverse`) to specify which column variable we want R to summarize for us. 

```{r, echo=TRUE}

bos_rents|>
  pull(rent) |>
  hist(breaks = 50)

```

_How would you describe the skew of our data based on this histogram?_

To look at a slightly less skewed subset of the data, let's filter the city-wide dataset to just 1 bedroom rental units and then display the histogram of asking rent.

```{r}
br_1 <- bos_rents |>
  filter(bedrooms == 1)

hist(br_1$rent)
```

# Summary Statistics

In this next section, we'll look more closely at the basic statistics of the `rent` variable. We'll begin with overall summary statistics that describe the central tendencies of the data (mean, median). We'll then move on to descriptive statistics and plots that help us describe the dispersion, or difference, present in our dataset.

The command `summary` provides us with the mean, median, 25th, and 75th quartiles, and minimum and maximum values...

```{r, echo=TRUE}
summary(bos_rents$rent)
```

...and for one-bedrooms in particular...

```{r, echo = TRUE}
summary(br_1$rent)
```

Finally, we can get narrow the range even more by looking at a particular neighborhood. Let's examine asking rents for 1-bedroom units in Chinatown...

```{r, echo = TRUE}
br_1_chinatown <- br_1 |>
  filter(neighborhood == "Chinatown")

summary(br_1_chinatown$rent)

# Note that we could do all of this, starting with our 
# entire dataset in one fell swoop like this...
# br_1_chinatown <- bos_rents |>
#   filter(neighborhood == "Chinatown" & bedrooms == 1)

# We could even do the variable selection in our pipeline
# and pass it directly to the summary function!
# bos_rents |>
#   filter(neighborhood == "Chinatown" & bedrooms == 1) |>
#   pull(rent) |>
#   summary()
```

The summary function can be graphically represented in the box plot. Box plots are one way that we can visualize the spread of our data.  Recall that the line in the center of the rectangle is the median, the rectange is the 'interquartile range'---i.e., the top is the 75th percentile and the bottom is the 25th---and the whiskers... Well, they're actually doing something a little fancy by default and not indicating the minimum to maximum, but rather trying to indicate a representative range while the plot still includes the outliers as points.


```{r}
boxplot(br_1$rent)
```
We can use one additional functionality to analyze the rent for each bedroom count. Let's pass in a function! The term on the left side (`bos_rents$rent`) is the values and the term on the right side (`bos_rents$bedrooms`) is a grouping variable. Each unique value of the latter will receive its own boxplot!

```{r}
boxplot(bos_rents$rent ~ bos_rents$bedrooms)
```

Finally, let's look at the cost of one-bedroom units by neighborhood. (We include the `las = 2` parameter, which simply rotates the labels vertically so that all will fit.

```{r}
boxplot(br_1$rent ~ br_1$neighborhood, las = 2)
```

## Measures of Central Tendency: Mean and Median

We can use R to calculate specific statistics, including the mean and median of rent, as follows.

**Note**: The `na.rm` argument in the function `mean` tells R how to handle missing values in the variable we are examining. When we set `na.rm=TRUE`, R removes missing values from its calculation.

We can use ``pull` to select the variable that we are analyzing: 

```{r, echo=TRUE}
bos_rents |>
  pull(rent) |>
  mean(na.rm=TRUE) 
```

Another way that we can do this is to use specific commands for `mean` and `median` :

```{r, echo=TRUE}
mean(bos_rents$rent, na.rm=TRUE)

median(bos_rents$rent, na.rm=TRUE)
```


## Quantiles

Let's say we wanted to know the rent amount at the 20th, 40th, 60th, and 80th percentile (the quintiles), to do so, we'd use the quantile function. 

```{r, echo=TRUE}
quantile(bos_rents$rent, probs=c(.2,.4,.6,.8))
```
These percentiles tell us what percentage of our data falls below each cut off point (e.g. the 20 percentile rent price indicates that 20% of rents in our data set are less than this value).


## Quantiles with Nested Function

You can combine the $\mathtt{seq}$ function that we saw in recitation 1 to make this a little easier:

```{r, echo=TRUE}
quantile(bos_rents$rent,probs=seq(.2,.8,by=.1))
```


## Measures of Dispersion

Measures of dispersion include: range, variance, and standard deviation. These statistics describe the spread or variability in the data. For data that are closely grouped together, these measures will have smaller values. Conversely, for data that are very spread out (or highly dispersed), these values will be greater.

We'll look at two such measures: range and standard deviation.

## Range

The range is the difference between the largest value and the smallest value present in your data. The base R function `range` will return the minimum and maximum of a vector in your dataset. 

To calculate the range directly in R, we can write an expression using the `min` and `max` commands in R, which identify the minimum and maximum values for a vector, respectively.

```{r}
range(bos_rents$rent) 

max(bos_rents$rent) - min(bos_rents$rent)
```
## Standard Deviation

Standard deviation is a measure of how data vary with respect to the mean. Let's do this for our one-bedroom subset.

R can calculate the standard deviation in your sample using the function `sd()`.

```{r}
sd(br_1$rent)
```

This is in units of the variable, so the standard deviation for a 1-bedroom apartment in Boston is \$`r sd(br_1$rent)`.

That's all for today!
