---
title: "R Markdown Template"
author: "Chenab Navalkha"
date: "2024-02-22"
output: html_document
---
^This is called a YAML Header - note that it includes title, author, and date for your file, as well as the type of file output you are specifying. 


# What is R Markdown?

R Markdown has a few more features than your typical R-script that allows you a little bit more say in how to organize your code when you're working in RStudio. Markdown works through "chunks" of code. Note that each chunk below is opened and closed with three backward apostrophes (```). 

You can run each chunk separately, or use the little arrows in the top right hand corner of each chunk to run all code above or below. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

One way to structure your R Markdown document is to create separate chunks for standardized parts of your workflow.

For example, we can start by loading in the packages we typically use:
```{r}
library(tidyverse)
library(xtable)
```

Then, we can set our working directory:
```{r}
setwd("/Users/chenabnavalkha/Desktop/QR/recitations/")
```

And then we can get going with the rest of our code...

```{r cars}
summary(cars) # cars is a dataset that is built in to R
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

You can customize which blocks of code and which output you want to include in your output by customizing the `knitr` options in your chunk header.

For example, if we want to include the code and the output in our export, we would open the chunk with the following parameters: 
```{r, include = FALSE, echo = FALSE}
# the "include" argument (TRUE/FALSE) determines whether code and results will appear in your finished file. Regardless of what value you set this to, R Markdown will still run the code.

# the "echo" argument (TRUE/FALSE) determines whether the results will appear in the finished file (if you use echo, the code will not appear in the finished file, even if you set this to "TRUE")
```

For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. 